---
title: "PCA Tutorial"
author: "Katherine Sheu"
date: "February 2, 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Put the title of your vignette here}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Here are some examples for using the package functions. 

PCA and PLSR scripts, with example input and output.

scripts:
glab.library/R/
	PCA_from_file.R
	varimax_from_file.R
	plot_pca.R
	PLSR_from_file.R
	plot_pls.R
input files:
glab.library/vignettes/PCA_tutorial/
	Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt
	human.info.rsem.expression.txt

Recommend running these scripts in RStudio, or a similar R GUI (graphical user interface)

Th PCA scripts are basically a wrapper for the R packace 'prcomp'.
	https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prcomp.html
PLSR scripts are a wrapper for R package 'pls'mixOmics'.
  https://cran.r-project.org/web/packages/mixOmics/index.html
  http://www.bioconductor.org/packages/release/bioc/html/mixOmics.html
  
#Open the five *.R files in RStudio. 

#'source' the five files to load the algorithms into memory. In more advanced applications they can be installed as packages. 

For help information on each parameter either (i) see the header portion of each .R file, or (ii) when installed as packages in R/RStudio, use help("PCA_from_file") or help('plot_pca') in R or RStudio. 


#set working directory.  Replace "/path/to/working/dir" with the actual working directory where the files are located
setwd("/path/to/working/dir")

```{r cars}

#execute the following commands from the console of the Rstudio GUI:


#read annotation file----
human.info = read.delim("human.info.rsem.expression.txt")
	
# for PCA run----
PCA_from_file("Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt", center = T, scale = F)
plot_pca("Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt",
    human.info$sample, human.info$type, labels = F, ellipse = T, PCx = "PC1", PCy = "PC2")

# for PLSR run----
PLSR_from_file("Beltran_2016_rsem_genes_upper_norm_counts_coding_log2.txt", human.info$sample, human.info$type,
    (ifelse(human.info$type=="NEPC", 1, 0)), comps = 5, scale = F)
plot_pls("Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_PLSR_Xscores.txt",
    human.info$sample, human.info$type, title = "PLSR", labels =F, PCx = "comp1", PCy = "comp2", ellipse = T, conf = 0.90)


# for varimax run ----
varimax_from_file("Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores.txt",
                  "Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_loadings.txt",
                  comp = 2, normalize = F)
plot_pca("Beltran_2016_rsem_genes_upper_norm_counts_coding_log2_prcomp_scores_VARIMAX.txt",
    human.info$sample, human.info$type, labels = F, ellipse = T)


```

